% this file is called up by thesis.tex
% content in this file will be fed into the main document

%: ----------------------- introduction file header -----------------------
\chapter{Introduction}\label{ch:chapter1}

% ----------------------------------------------------------------------
%: ----------------------- introduction content ----------------------- 
% ----------------------------------------------------------------------

Ageing populations experience substantial changes across physical, nutritional, cognitive, and psychosocial domains, each shaping vulnerability to chronic disease and functional decline~\cite{davisComparisonGaitSpeed2021, mooreCountingTimeLived2017, huangSexDifferencesAssociations2020}. Understanding how these multidomain factors interact over time is essential for early risk identification.

The financial and societal burden is substantial. In Europe, poor diet is a major risk factor for non communicable diseases, contributing billions in healthcare costs annually~\cite{WhoNutritionReport}. In Ireland, malnutrition alone is estimated to cost over €1.4 billion per year~\cite{RICE2012}. In response to these challenges, international public health frameworks, including the \gls{WHO} \emph{Global Action Plan} on noncommunicable diseases and Ireland's \emph{Healthy Ireland} framework, highlight nutrition and lifestyle factors as central components of disease prevention strategies.

European regulatory frameworks such as the \gls{GDPR} and proposed \gls{EU} \gls{AI} Act emphasise transparency and highlight the need for interpretable \gls{ML} models in healthcare.

Traditional risk prediction models have largely relied on clinical biomarkers such as \gls{BP} and lipid profiles. Yet, growing evidence shows that broader predictors, including nutrition~\cite{murphyPlasmaLuteinZeaxanthin2023}, \gls{PA}~\cite{lairdPhysicalActivityDose2023}, \gls{MH}~\cite{mcdowellAssociationsPhysicalActivity2018}, and \gls{SOC}~\cite{mccrorySocialDisadvantageSocial2016} improve predictive performance. Estimating the likelihood of developing a condition given certain risk factors helps quantify the influence of these predictors in clinical decision making. 

\gls{TILDA}\footnote{\gls{TILDA}'s dataset are pseudonymised and accessed under license via \gls{ISSDA}. Use adheres to \gls{UL} S\&E ethics approval and data retention requirements.} provides a national representative cohort of adults aged $\texttt{50+}$ in Ireland, surveyed repeatedly across multiple domains. An externally produced \gls{RAND} Harmonised \gls{TILDA} dataset (available for waves $1-2$) was also inspected and prepared during data processing, its role and potential for cross study replication are discussed in Chapter~\ref{ch:chapter3} and future work (Chapter~\ref{ch:chapter6}).

Despite these advantages, existing \gls{ML} research on chronic disease prediction faces persistent limitations: over reliance on clinical predictors, limited incorporation of lifestyle and \gls{SOC} variables, predominantly cross sectional study designs that fail to capture trajectories, and a lack of interpretability consistent with emerging \gls{AI} governance standards.  

This thesis addresses these gaps by leveraging longitudinal \gls{TILDA} data across multiple domains, using interpretable \gls{ML} methods to combine modifiable health behaviours, psychosocial factors, and established clinical measures to enhance prediction of ageing related outcomes.
% ----------------------------------------------------------------------
% Research Problem
\section{Research problem and motivation} 

Despite increasing research and clinical interest in lifestyle, nutrition factors, and advances in \gls{ML} methods, approaches to chronic disease prediction in older adults remain limited in several ways.

\begin{itemize}
    \item \textbf{Limited Integration of Multidomain Predictors:} Existing models focus primarily on clinical measurements and often neglect modifiable behavioural, dietary, and psychosocial factors~\cite{murphyPlasmaLuteinZeaxanthin2023, lairdPhysicalActivityDepression2023}.
    \item \textbf{Underuse of Longitudinal Dynamics:} Few studies leverage repeated measures to capture developments of risk, such as \gls{FR} progression~\cite{moguilnerImportanceAgePrediction2021}.
    Probabilistic reasoning can help estimate the likelihood of developing a condition at a future time given earlier measurements, for example, calculating the probability $P(\text{\gls{FR} at wave 5} \mid \text{\gls{UGS} at wave 1}, \text{Nutrition at wave 2})$.

    This approach helps quantify how early indicators influence later outcomes and supports better feature selection and timing of interventions in predictive models.
    \item \textbf{Lack of Explainability:}Many \gls{ML} models are not interpretable, limiting clinical adoption and failing to meet emerging regulatory requirements for trustworthy \gls{AI} in healthcare~\cite{donoghueUsingConditionalInference2023}.
\end{itemize}
Addressing these challenges is essential to improve predictive accuracy, support evidence based interventions, and ensure alignment with evolving policy and regulatory priorities.

% ----------------------------------------------------------------------
% Research Aims and goals
\section{Research aims and objectives}

The goal of this thesis is to develop and validate \gls{ML} models that integrate nutrition, lifestyle, and \gls{MH} features with clinical data to improve prediction of chronic disease and ageing related outcomes in a representative cohort of older adults. This goal will be achieved through the following objectives:

\begin{itemize}
  \item[-] Review existing applications in \gls{ML} for chronic disease prediction in ageing populations, with emphasis on gaps in lifestyle and nutrition integration.
  \item[-] Develop a reproducible pipeline to preprocess and align \gls{TILDA} waves 1--5, and engineer multidomain features (nutrition, activity, mobility, \gls{MH}, clinical indicators).
  \item[-] Develop and compare a linear baseline (\gls{LR}) against tree ensemble and neural network families (\gls{XGB}, \gls{RF}, \gls{MLP}) across single wave and pooled longitudinal designs.
  \item[-] Evaluate models using predictive classification performance metrics including \gls{AUC}, calibration and error measures (accuracy, precision, \gls{recall}, \gls{F1}), perform ablation analyses to assess the incremental value of lifestyle and nutrition predictors.
  \item[-] Identify the most informative predictors of chronic disease onset and assess whether domain specific features beyond the core health burden set contribute independent predictive value.
  \item[-] Document ethical, governance, and reproducibility practices for handling longitudinal pseudonymised health data.
\end{itemize}

% ----------------------------------------------------------------------
% Contributions
\section{Contributions of the research}

This research makes contributions at three levels:

\begin{itemize}
  \item \textbf{Empirical:} Demonstrates that predictive signal for chronic disease onset in this cohort is concentrated in a compact set of health burden indicators (medication use, self-rated health, pain, functional limitation and other \gls{MH} features), that pooled longitudinal modelling substantially improves both performance and stability over a single-wave baseline. % A bit cliche i think we dont have to explecity mention this becaise then teh nrration becomes ot evdienty taht this was fitted to jusitfy our chapter 5 hahaha lets make it natrual right i think we shoudl drop the statemnt of single wave vs pooled longitudinal because it is a bit too much of a spoiler for the results chapter, we can just say that we demonstrate the value of longitudinal modelling and then in the results chapter we can show the comparison between single wave and pooled longitudinal and then we can discuss the implications of that in the discussion chapter, i think that would be a more natural way to present the results rather than giving away the punchline in the introduction. Again try keep things as short as possible
  \item \textbf{Methodological:} Provides a reproducible pipeline for aligning longitudinal \gls{TILDA} data across domains and integrating diverse features into \gls{ML} workflows.  
  \item \textbf{Clinical:} Highlights actionable modifiable factors alongside traditional clinical indicators for targeted interventions and policy. 
\end{itemize}

% ----------------------------------------------------------------------
% Outline for the rest of the thesis
\section{Thesis outline}

\emph{Chapter}~\ref{ch:chapter2} presents a literature review on \gls{ML} prediction of chronic disease in ageing populations, focusing on lifestyle, nutrition predictors and methodological considerations.
\emph{Chapter}~\ref{ch:chapter3} describes the data preparation process, including the \gls{TILDA} study design, data collection, thematic domain structure, and the manual variable filtering workflow.
In \emph{Chapter}~\ref{ch:chapter4}, the automated cleaning pipeline is presented, including the variable scenario detection system, the deterministic six-step cleaning sequence, and validation of the pipeline against established benchmarks.
\emph{Chapter 5} presents the empirical evaluation, covering performance across three modelling designs, disease specific prediction, predictor add-on analysis, and an interpretation of the health burden correlation structure underlying the results.
\emph{Chapter 6} discusses the implications and future directions of the research.

% ---------------------------------------------------------------------------
% ----------------------- end of thesis sub-document ------------------------
% ---------------------------------------------------------------------------