% Thesis Abstract -----------------------------------------------------
% BMRC style

%\begin{abstractslong}    %uncommenting this line, gives a different abstract heading
\begin{abstracts}        %this creates the heading for the abstract page.

\textbf{Background:} Chronic disease and functional decline are major health challenges for older adults, yet how well self-reported health burden can predict disease onset across longitudinal waves remains underexplored.

\textbf{Methods:} Using five waves of data (2009--2019), a longitudinal machine learning framework evaluated two tasks: predicting any chronic disease onset across three designs (single wave, two pooled multi-wave), and 24 disease specific outcomes. Four model families were compared via cross-validation (CV) area under the curve (AUC), with predictor contributions assessed by bootstrapped add-on analysis.

% ToDo: Should we replace the word variable with features or predcitors? we also use the word predictor one sentence prior but lets use what is correct mentioning variables seemed off to me for an overall satistical mention. Sp two things: recheck the use of predictors in the sentence of '20 most important predicors' and variables in 'addon variables showed'
\textbf{Results:} Pooled multi-wave modelling improved by $22\%$ over the single wave baseline ($0.718 \pm 0.006$ vs $0.589 \pm 0.072$), with cross-fold variance decreasing more than tenfold. Disease specific models achieved test performance up to $0.721$ for functional outcomes. Reducing to the 20 most important predictors caused no measurable performance loss, while cognitive, psychosocial, nutritional, and behavioural add-on variables showed no statistically significant improvement.

\textbf{Conclusions:} Among community dwelling older adults, chronic disease risk prediction was primarily driven by self-reported burden indicators, particularly medication use, pain, and self-rated health rather than domain specific biomarker or lifestyle profiles. These findings suggest that publicly available survey data can support meaningful risk stratification with a limited feature set.

\textbf{Keywords:} ageing, \glsentryfull{ML}, predictive modelling, chronic disease, functional decline, health burden, \glsentryfull{TILDA}, \glsentryfull{XGB}.

\end{abstracts}
%\end{abstractlongs}

% ---------------------------------------------------------------------------
% ----------------------- end of thesis sub-document ------------------------
% ---------------------------------------------------------------------------