% this file is called up by thesis.tex
% content in this file will be fed into the main document

\chapter{Conclusion}\label{ch:chapter6}


% ----------------------- paths to graphics ------------------------

% change according to folder and file names
\ifpdf
    \graphicspath{{6_conclusions/figures/PNG/}{6_conclusions/figures/PDF/}{6_conclusions/figures/}}
\else
    \graphicspath{{6_conclusions/figures/EPS/}{6_conclusions/figures/}}
\fi


% ----------------------- contents from here ------------------------


\section{Summary}
This thesis developed and evaluated a longitudinal \gls{ML} pipeline for predicting chronic disease onset in older adults using all five waves of \gls{TILDA} data. Chapter~\ref{ch:chapter3} described the study population and manual variable screening workflow, which reduced over 5,000 raw variables to a structured, domain organised dataset. Chapter~\ref{ch:chapter4} presented the deterministic six step cleaning pipeline and validated the transformed features against published benchmarks. Chapter~\ref{ch:chapter5} evaluated four model families across three modelling designs and 24 disease specific outcomes, finding that pooled multi-wave designs substantially outperformed the single-wave baseline, that predictive signal concentrated in a small cluster of health burden indicators, and that no domain specific add-on feature contributed statistically significant independent predictive value. 
Code and pipeline scripts supporting all analyses are available via the repository referenced in Section~\ref{sc:contributions}.

% ---------------------------------------------------------------------------
\section{Conclusions}
Longitudinal pooling works. Expanding beyond the core health burden predictor set adds no statistically significant independent gain. The performance and stability gains documented in Chapter~\ref{ch:chapter5} reflect a structural alignment between predictor horizon and outcome definition rather than simply sample size. Predicting clinically coded disease eight years out is a fundamentally harder task than predicting self-reported health burden over two-year transitions, and matching these two layers drives the gain. Restricting the feature set to 20 predictors caused no further loss. The domain specific add-on analysis confirmed that psychosocial, nutritional, lifestyle, and sociodemographic features showed no statistically significant independent contribution beyond the core health burden cluster, consistent across all four model families and all three design configurations.

Chapter~\ref{ch:chapter2} framed prediction as estimating $P(\text{disease} \mid X)$, where the marginal gain from any added feature depends on how much independent information it carries. The domain counts make this concrete. Of the $3{,}133$ retained variables (Table~\ref{tab:feature_groups}), sociodemographic context contributed $844$ variables, psychological and cognitive measures $291$, and nutritional biomarkers $22$, none of which produced a single top-20 predictor. Lifestyle behaviours ($120$ variables) contributed one item: flu vaccination, which functions in this context as a proxy for primary care engagement and health system contact rather than a modifiable behaviour. Across all four of these domains combined, predictive signal was effectively absent. The features that drove nearly all discriminative signal came from physical health ($853$ variables) and functional status ($174$ variables). Even then, just 20 of those $1{,}027$ variables sufficed. Self-rated health, medication burden, and pain appear to act as integrative markers that already absorb the signal from the other domains. Adding $1{,}277$ variables across four domains that are genuinely important in ageing does not move the posterior.

The validation in Chapter~\ref{ch:chapter4} adds a dimension often missing from published work. Encoding decisions can materially affect model performance in health survey data. Recoding six frailty index items to respect published ordinal deficit scoring (where each item must remain bounded between $0$ and $1$ to preserve clinical meaning)~\cite{searleStandardProcedureCreating2008} improved \gls{AUC} from $0.591$ to $0.636$ under the biologically informed proxy. This $7.6\%$ gain is attributable entirely to encoding, with no change in the feature set or model.

The three design structure forms a natural test of what matters. Adding temporal structure (D1 to D2) improved performance substantially. Reducing the feature set from $261$ to $20$ (D2 to D3) produced negligible further change, with a test \gls{AUC} difference of only $0.002$ for \gls{XGB}, confirming that the signal is compact and concentrated. The designs differ in their feature selection parameters, with D1 using $k=20$ from approximately $570$ available features and D2 using $k=100$ from approximately $261$ features after missingness filtering. These thresholds were determined through systematic experiments to be appropriate for the respective sample sizes and feature pools, and are not fixed choices applied uniformly across designs. D3 resolves any comparability concern directly. Applying the top-20 feature set to the full D2 cohort recovered D2 level performance within $\Delta$\gls{AUC} $= 0.002$, confirming that the cross-design performance gap reflects longitudinal structure rather than the number of features retained. Hyperparameter optimisation in D3 produced similarly negligible gains, consistent with a dataset where the available signal is already well captured by the core health burden features. 

Track~B reinforces the pattern. Outcomes with sufficient incidence and consistent wave measurement generalised well, while outcomes depending on sparse events or indirect proxy targets did not, regardless of algorithm or feature strategy. The implication for \gls{ML} on community health survey data is direct. Prediction is bounded by the depth of the data, not the complexity of the model. Substantive gains beyond the current ceiling require the linked records held under elevated access (mortality registers, primary care histories, and biomarker panels), as outlined in Section~\ref{sc:future_work}.

\subsection{Data availability} 
This analysis was conducted using the publicly accessible \gls{TILDA} release (waves~1--5), a pseudonymised dataset covering self-reported health, physical, cognitive, nutritional, and behavioural variables, together with clinical coded disease indicators from wave~2 onwards. The datasets are accessible through the \gls{ISSDA} at \url{https://tilda.tcd.ie/data/accessing-data/}.

Several data sources relevant to chronic disease prediction are not included in the public release. These require on-site access at \gls{TILDA}'s Trinity College Dublin facility or remote access through the \gls{TILDA} VISTA trusted research environment, and include a dedicated biomarkers dataset, the wave~3 \gls{MRI} dataset, an epigenetics dataset, linked primary care records, and linked mortality records. Access is subject to formal application to the \gls{TILDA} data access committee and operates on a cost recovery basis where institutional funding is available. Future work drawing on these sources would require updated ethics approval, \gls{GDPR} training certification and a signed data use agreement with \gls{TILDA}.

\subsection{Ethics statement}  
The \gls{TILDA} study was reviewed and approved by the Faculty of Health Sciences Research Ethics Committee at Trinity College Dublin, and all participants provided written informed consent to their data being used for research purposes. The secondary analysis conducted in this thesis was approved by the University of Limerick Science and Engineering Research Ethics Committee. All data used were publicly released, pseudonymised datasets accessed through \gls{ISSDA}. No direct participant contact or identifiable personal data were involved in this analysis.

% ---------------------------------------------------------------------------
\section{Contributions}
This research makes contributions at three levels. \textbf{Empirically}, it demonstrates that pooled longitudinal modelling substantially improves predictive performance and stability for chronic disease onset in the \gls{TILDA} cohort, and identifies a compact cluster of self-reported health burden indicators as the primary driver of that signal. Domain features outside this core set showed no statistically significant independent predictive value, a result consistent across all model families and designs. \textbf{Methodologically}, a transparent, reproducible preprocessing pipeline for aligning heterogeneous longitudinal health survey data is documented in full, addressing a gap rarely covered in published \gls{ML} literature. \textbf{Clinically}, the results provide a data driven basis for risk stratification using publicly available survey data, and identify the conditions under which more complex predictor sets or model families are unlikely to improve performance without access to richer linked data.

The results in Table~\ref{tab:lr_table} provide a concrete benchmark. Test \gls{AUC} values in D2 and D3 ($0.71$--$0.72$) sit at the upper end of the $0.64$--$0.71$ range documented in comparable longitudinal prediction studies.~\cite{donoghueUsingConditionalInference2023} report \gls{AUC} $\leq 0.69$ for falls prediction in \gls{TILDA} (waves~1--3), and~\cite{zhangMachineLearningModels2025} report $0.64$--$0.71$ for longitudinal mental health disorder prediction. The present results are comparable with, and in some cases above, these benchmarks while targeting a broader outcome using only publicly accessible data. The single-wave baseline (D1: $0.53$--$0.58$) confirms that the gain is driven by the pooled longitudinal design, not algorithm choice. All four model families cluster within $0.001$ \gls{AUC} in D3, which reinforces this. The gap to~\cite{xuTailoredMachineLearning2023} ($0.854$) reflects task specificity. Predicting diabetes from 105 curated predictors drawn from over 40,000 variables via a stacked ensemble is not comparable to predicting general chronic disease onset from public survey features.

% ---------------------------------------------------------------------------
\section{Future work}\label{sc:future_work}
The recently released sixth wave of \gls{TILDA} data (December 2025), together with the Intellectual Disability Supplement and the COVID-19 sub study, would extend the longitudinal coverage and introduce health trajectory data spanning a pandemic period, likely altering self-reported burden profiles in meaningful ways. Cross cohort validation using the \gls{RAND} harmonised \gls{TILDA} dataset, which maps variables to international standards compatible with cohorts such as \gls{HRS} and \gls{SHARE}, would test how well the identified health burden cluster generalises beyond the Irish context. Finally, access to the restricted biomarker panels, primary care records, and linked mortality data held within \gls{TILDA}'s secure research environment would allow future work to test whether the signal limits identified here reflect the nature of chronic disease risk itself or the constraints of the publicly available dataset.

The encoding challenges encountered in preparing the \gls{TILDA} datasets point to a wider gap in the \gls{ML} literature. Standardised preprocessing protocols and auditable encoding indices for longitudinal health survey data do not yet exist as a shared resource. Establishing reproducible, domain aware transformation workflows would reduce the risk that performance differences across studies reflect encoding choices rather than genuine signal, supporting more reliable cross study comparisons in health data science.


% ---------------------------------------------------------------------------
% ----------------------- end of thesis sub-document ------------------------
% ---------------------------------------------------------------------------