% this file is called up by thesis.tex
% content in this file will be fed into the main document

%: ----------------------- name of chapter  -------------------------

\chapter{Literature Review} % top level followed by section, subsection

%: ----------------------- paths to graphics ------------------------

% change according to folder and file names
% \ifpdf
%     \graphicspath{{2_chapter2/figures/PNG/}{2_chapter2/figures/PDF/}{2_chapter2/figures/}}
% \else
%     \graphicspath{{2_chapter2/figures/EPS/}{2_chapter2/figures/}}
% \fi

%: ----------------------- contents from here ------------------------

This chapter reviews literature on ageing, chronic disease, and predictive modelling using \gls{TILDA}. 
The focus is on three domains central to later life health: frailty and physical function, mental health, and nutrition and biological ageing. 
Chronic conditions including multimorbidity, frailty, and mental health difficulties are leading causes of morbidity and reduced quality of life in older adults, often overlapping to accelerate functional decline. 
In line with the \gls{WHO} healthy ageing framework, research using \gls{TILDA} highlights frailty, mental health, and nutrition as interconnected determinants of later life outcomes. 
The following subsections examine each domain in turn.\label{sec:ChronicDiseaseAndAgeing}

% ----------------------------------------------------------------------
% Frailty and Physical Function
\section{Frailty and Physical Function}

Frailty is a reflection of higher vulnerability and declining physiological reserve. 
In \gls{TILDA}, it is captured through both the frailty phenotype and the frailty index, with gait speed, grip strength, and mobility decline serving as practical predictors. 
These measures strongly predict hospitalisation and mortality.

\subsection{Probabilistic Interpretation of Frailty}
One can argue that from a probabilistic perspective, the chance of an adverse outcome (such as mortality, $Y$) given observed features (such as frailty markers, $X$) as a conditional probability $P(Y \mid X)$. 
This allows us to deduce which frailty features are most predictive of adverse outcomes. Features may be independent, dependent, or mutually exclusive. 

The presence or absence of metabolic syndrome represents a mutually exclusive event that can inform risk models.
\begin{equation*}
P(X_i \cap X_j) = P(X_i) P(X_j)
\end{equation*}
for independent features $X_i$ and $X_j$, guiding feature selection and reducing redundancy in model inputs.
This reflects the likelihood that an individual with certain frailty characteristics will experience a particular outcome.

Evidence suggests that frailty is dynamic, not irreversible. 
For example, multi state modelling showed that many prefrail adults transition back to non frail, although established frailty strongly predicts mortality \cite{romero-ortunoTransitionsFrailtyPhenotype2021}. 
Metabolic syndrome, on the other hand, speeds up the onset of frailty, which suggests a metabolic pathway of vulnerability. \cite{mccarthyAssociationMetabolicSyndrome2023}. 
The presence of metabolic syndrome modifies the conditional probability $P(\text{Frailty} \mid \text{Metabolic Syndrome})$ and may increase the likelihood of adverse outcomes over time. 
Another detail is that age matters:  obesity predicts mobility decline earlier in life,  whereas later in life, traits that protect against this, including grip strength, become more significant \cite{huangAgeModifyAssociations2019}. 

These findings partly conflict: frailty appears both reversible and progressive, depending on risk context. 
This complexity highlights the need for probabilistic models that can update beliefs about risk as new features are observed, as formalised by Bayes' theorem:
\begin{equation*}
P(Y \mid X) = \frac{P(X \mid Y) P(Y)}{P(X)}
\end{equation*}
where $P(Y \mid X)$ is the posterior probability of outcome $Y$ given features $X$. 

\subsection{Bayesian Updating of Frailty Risk}
This enables us to incorporate prior knowledge and observed data, improving the interpretability of feature outcome relationships in ageing.
Here, $P(Y \mid X)$ is the posterior probability = ( Likelihood * Prior / Evidence ), representing the updated belief about the outcome after observing features. 
$P(Y)$ is the prior probability, reflecting baseline risk. 
$P(X \mid Y)$ is the likelihood, quantifying the probability of observed features given the outcome, and 
$P(X)$ is the evidence (the overall probability of the features). 
This framework allows us to update predictions as new data is observed, improving interpretability in feature outcome relationships.

Limitations remain. Causal inference is limited by observational designs, and predictors are frequently examined separately, ignoring interactions. 
Most frailty models have only been with the \gls{TILDA} dataset and with this present study focuses on the same \gls{TILDA} but with an expansion on the harmonised set to explore more longitudinal prediction within this context. 
These gaps point to the need for multidimensional approaches. 
Traditional regression models may not fully capture age dependent interactions, suggesting a role for more flexible methods such as machine learning techniques and models.

% ----------------------------------------------------------------------
% Mental Health in Older Adults
\section{Mental Health in Older Adults}

Mental health, including depression, anxiety, loneliness, and social isolation, strongly shapes quality of life. 
Evidence from \gls{TILDA} shows strong links to both physical and social factors. \cite{santiniProtectivePropertiesActBelongCommit2017} found that participation in physical, social, and cognitive activities summarised by the Act-Belong-Commit (ABC) indicator was associated with lower depression and anxiety. 
This suggests social integration and lifestyle protect mental wellbeing. However, \cite{juolaO117NursingStaff2014} reported that self reported measures may exaggerate such associations..., raising concerns of bias. 
These two findings conflict: the effect of social engagement is strong when self reported, weaker when objectively assessed.

Objective data strengthen the case for activity. Accelerometer based measures show a dose response link between moderate to vigorous activity and lower depressive symptoms \cite{lairdPhysicalActivityDose2023}. 
Yet reverse pathways are also evident: incident falls and functional decline predict later depression, refereed by \gls{FOF} \cite{jacobFallsIncreaseRisk2023}. 
Together, these results reveal a bidirectional relationship, mental health both influences and is influenced by physical decline.
Most studies rely on broad self report scales such as \gls{CESD} or \gls{HADS}, with little biomarker or clinical validation. 

This literature suggests mental health in ageing cannot be modelled as a one way effect. 
It requires longitudinal, multidimensional approaches that can capture mutual pathways, an area where machine learning is well suited especially when combined with probabilistic models that estimate the likelihood and posterior probability of mental health outcomes given multiple interacting features.

For example, we can estimate:
\begin{equation*}
P(\text{Depression} \mid \text{Activity, Falls, Social Engagement}),
\end{equation*}
quantifying how these interacting features together influence mental health outcomes. 
Probabilistic modelling helps determine which features have the strongest effect and which may be redundant for predictive purposes.

% ----------------------------------------------------------------------
% Nutrition and Biological Ageing
\section{Nutrition and Biological Ageing}
Nutrition is increasingly recognised as a modifiable factor in ageing and frailty. 
Biomarkers like vitamin D, lutein, and zeaxanthin have been linked to physical and cognitive outcomes in \gls{TILDA}.
\cite{murphyPlasmaLuteinZeaxanthin2023} showed that higher plasma lutein and zeaxanthin dietary carotenoids found in leafy vegetables were linked to slower frailty progression. 
By contrast, \cite{mccarthyAssociationVitaminDeficiency2022} found vitamin D deficiency predicted diabetes and musculoskeletal decline, especially among adults with low sunlight exposure. 
These studies point to different nutritional mechanisms: antioxidant protection versus metabolic regulation.


\subsection{Bayesian Perspective on Nutritional Biomarkers}
The effect of nutrition on ageing outcomes can be formalised probabilistically using Bayes theorem:
\begin{equation*}
P(\text{Outcome} \mid \text{Nutritional Status}) = \frac{P(\text{Nutritional Status} \mid \text{Outcome}) P(\text{Outcome})}{P(\text{Nutritional Status})},
\end{equation*}
where the likelihood function $P(\text{Nutritional Status} \mid \text{Outcome})$ quantifies how observed nutritional biomarkers relate to health states, and the posterior $P(\text{Outcome} \mid \text{Nutritional Status})$ captures updated beliefs about outcomes, including uncertainty in small sample studies.
This helps identify which biomarkers are most informative for predicting ageing outcomes which supports feature selection in \gls{ML} models.

\subsection{Contextual and Integrative Considerations}
Malnutrition also reflects social and functional risks. \cite{bardonPredictorsIncidentMalnutrition2020} found that hospitalisation, functional loss, and poor social support increased malnutrition risk, with notable sex differences. 
This highlights a limitation in much nutritional research: biological markers are studied without integrating social context.
The difference between biological and chronological age is a major point of debate. 
Biomarker based metrics that better capture these cumulative exposures, like frailty indices or epigenetic clocks, frequently vary from chronological age. 
McCarthy and Azizi link metabolic syndrome and nutritional deficits to accelerated biological ageing \cite{mccarthyMetabolicSyndromeAccelerates2023,aziziEvidenceAssociationAllostatic2024}. 
Although the majority of the evidence is cross sectional and has little ability to capture changes over time, these indicators offer a more precise prediction of vulnerability. 
Reliability is also decreased in biomarker research with small sample sizes, which can be formalised through the likelihood function $P(\text{Data} \mid \theta)$ and its impact on posterior uncertainty in Bayesian inference.

Taken together, nutrition interacts with biological, metabolic, and social systems. 
Isolating single nutrients risks oversimplification. Integrative modelling that combines biomarkers, clinical factors, and social determinants supported by machine learning pipelines offers a more promising route. 
Models that estimate the joint distribution of features and outcomes can better capture the complex dependencies relevant to biological ageing.

% ----------------------------------------------------------------------
% Machine Learning in Ageing and Chronic Disease
\section{Machine Learning in Ageing and Chronic Disease}
\gls{ML} is increasingly applied in ageing research because it can capture non linear and multidimensional relationships that traditional regression models, such as logistic regression,often miss \cite{xuTailoredMachineLearning2023, boyleBrainpredictedAgeDifference2021}. 

\subsection{Feature Sets in ML Studies}
Studies using \gls{ML} in ageing research include many types of predictors: clinical measures, sociodemographic factors, lifestyle habits, cognitive tests, and psychosocial variables. 
A recurring pattern, however, is that models often underline easily measured physical health data, such as comorbidity counts, mobility scores, or vital signs, while more complex but important predictors, including nutrition and social factors, are underrepresented \cite{zarudskyP277OneleggedBalance2014}. 
This imbalance may limit the ability of models to capture the full complexity of ageing, especially considering that mental health and nutrition exert independent and interacting effects on later life outcomes see (\hyperref[sec:ChronicDiseaseAndAgeing]{here}). 
Models that focus only on readily available clinical data risk reproducing the same biases seen in traditional regression approaches.

\subsection{ML Models Used}
The methodological approaches used in ageing research span both traditional and advanced models. 
Logistic regression and Markov models remain common baselines due to their interpretability and simplicity, yet they are often unable to capture non linearities or higher order interactions. 
More recent work has shifted towards ensemble methods such as random forests (\gls{RF}) and gradient boosting machines (\gls{GBM}), as well as deep neural networks (\gls{DNN}), which generally achieve higher predictive accuracy.

Similarly, \cite{matthewsPredictingMortalityIrish2022} evaluated mortality prediction using a standard logistic regression model with demographic, clinical, and functional predictors, achieving an \gls{AUC} of 0.78, whereas, ensemble methods in \cite{xuTailoredMachineLearning2023} reached 0.854 for diabetes prediction, showing how flexible, non linear models can better capture complex interactions among ageing related predictors.

\subsection{Performance and Validation}
Predictive performance across studies is generally moderate to strong, with \gls{AUC} values between 0.70 and 0.90 for outcomes like diabetes, gait speed, depression, and mortality. 
This indicates that models can distinguish reasonably well between individuals who will or will not experience an outcome. 

However, there are important limitations. Many studies rely only on internal validation, testing models on the same data used for training. 
This can inflate performance estimates and reduce confidence in applying the model to new populations. 
To add, Most studies are cross sectional, so causal or temporal effects cannot be assessed. 
When combined with the underuse of nutrition and social predictors, these issues indicate that current models may not fully represent the multidimensional nature of ageing and could produce context specific results rather than generalisable predictions \cite{zarudskyP277OneleggedBalance2014}.

\subsection{Explainability in ML for Ageing}
As \gls{ML} models become more complex, interpretability emerges as a critical challenge. Explainability tools like \gls{SHAP} can highlight feature importance, but their use is inconsistent, particularly in deep learning models \cite{xuTailoredMachineLearning2023,davisLinearRegressionbasedMachine2021,davisComparisonGaitSpeed2021}. 
These methods improve transparency and can highlight unexpected drivers of outcomes, such as psychosocial variables being stronger predictors than biomedical ones, yet black box concerns remain, especially for deep learning models, where feature interactions are difficult to track.
For ageing research, where clinical adoption requires both accuracy and trust, explainability is not optional but essential. 
Current efforts show promise but remain fragmented, with no consensus on best practice. 
This gap links directly to the broader methodological limits discussed in the next section on gaps in the literature.

% ---------------------------------------------------------------------------
% Summary of Literature - Table
\begin{table}[htbp]
\centering
\caption{Summary of selected studies on ageing using \gls{TILDA}}
\label{tab:lr_table}
\small
\resizebox{\textwidth}{!}{
\begin{tabular}{p{2.5cm} p{2.5cm} p{2cm} p{4cm} p{4cm} p{4cm} p{4cm}}
\hline
\textbf{Theme} & \textbf{Author/year} & \textbf{Dataset} & \textbf{Variables/features} & \textbf{Methods/model type} & \textbf{Key Findings/outcome} & \textbf{Limitations} \\
\hline
Frailty & Romero-Ortuno (2021) \cite{romero-ortunoTransitionsFrailtyPhenotype2021} & Wave 1-5 & Frailty state, Frailty phenotype (FP) components (Exhaustion(EX), Unexplained weight loss(WL), Weakness(WK), Slowness(SL), Low physical activity(LA)) & Multi state Markov model, longitudinal 8 year follow up & 32\% of pre frail to non frail; frailty strongly predicted 31\% mortality & Self report bias; observational design \\
\hline
Falls + \gls{ML} & Donoghue (2023) \cite{donoghueUsingConditionalInference2023} & Wave 1-3 & Gait, mobility, medications, history of falls & Conditional inference forest; 5-fold \gls{CV}; undersampling for imbalance & Accuracy up to 69.7\%; for adults aged 50-64.& \gls{AUROC} values were $<= 0.69$ and \gls{PRAUC} $<= 0.39$ indicating low predictive accuracy; no external validation; cross sectional; Class imbalance \\
\hline
\gls{FOF} & Zarudsky (2014) \cite{zarudskyP277OneleggedBalance2014} & Wave 1 & Age, sex, fall history & Cross sectional prevalence analysis & 23.3\% prevalence; higher in women; 70\% with \gls{FOF} had no recent falls in the last 12 months & Self reported; cross sectional \\
\hline
Functional Mobility & Huang (2019) \cite{huangAgeModifyAssociations2019} & Wave 1-2 & Measured by \gls{TUG}, grip strength, physical activity, vision, \gls{BMI}, age & Linear regression, stratified analyses & Age modifies effects; \gls{PA} and grip protect mobility; vision benefit only in middle aged (Beta = -0.032, P = 0.002) & Observational; 2-year follow-up \\
\hline
\multirow{2}{*}{Depression} 
& Jacob (2023) \cite{jacobFallsIncreaseRisk2023} & Wave 1-2 & Falls, \gls{FOF}, \gls{HADS}-A, \gls{CESD} & Prospective logistic regression & Falls predicted incident anxiety (\gls{OR}=1.58) and depression (\gls{OR}=1.43) & Self-reported falls; short follow up; mediation effect partially explored \\
& Laird (2023) \cite{lairdPhysicalActivityDepression2023} & Wave 1-5 & \gls{MVPA}, walking, sociodemographics & Longitudinal regression; mixed model & Higher physical activity reduced depressive symptoms; dose response effect & Self-reported \gls{PA}; observational; limited causal inference \\
\hline
\multirow{2}{*}{Nutrition} 
& Murphy (2023) \cite{murphyPlasmaLuteinZeaxanthin2023} & Wave 1-5 & Plasma lutein, zeaxanthin & Linear/ordinal logistic regression & Higher carotenoids linked to slower frailty progression Lower odds of becoming frail (ORs = 0.57-0.89) & Observational; not predictive of longitudinal muscle changes \\
& Bardon (2020) \cite{bardonPredictorsIncidentMalnutrition2020} & Wave 1-2 & Hospitalisation, functional status, social support, sex, falls & Logistic regression; sex-stratified analyses & Functional decline, hospitalisation, poor support increased malnutrition risk; sex differences & Observational(short follow up (2 years)); limited generalisability >65 years, self reported\\
\hline
\multirow{3}{*}{\gls{ML}} 
& Xu (2023) \cite{xuTailoredMachineLearning2023} & Wave 1-3 & 105 predictors from 40,000+ variables (social, clinical, mental, family) & Stacked ensemble: \gls{RF}, XRT, GLM, \gls{GBM}, \gls{DNN}; \gls{SHAP} explainability & AUC=0.854 for 2 year diabetes prediction; top predictors LDL, cirrhosis, sex & No external validation; short follow up; ensemble complexity limits deployment \\
& Davis (2021) \cite{davisComparisonGaitSpeed2021} & Wave 3 & Age, grip strength, chair stands, \gls{BMI}, cognitive tests & Histogram gradient boosting regression; \gls{SHAP} & r$^2$  test: \gls{UGS}$\approx$0.41, \gls{MGS}$\approx$0.46, GSR$\approx$0.21 & Modest predictive power; cross sectional; no external validation \\
& Zhang (2025) \cite{zhangMachineLearningModels2025} & Longitudinal IMAGEN cohort & Psychopathology, personality, cognition, substance use, environment & Regularized logistic regression; diagnostic & Longitudinal AUC: ED=0.71, MDD=0.64, AUD=0.67; shared transdiagnostic predictors & Limited age range; no external replication \\
& Boyle (2021) \cite{boyleBrainpredictedAgeDifference2021} & Wave 3 & \gls{brainPAD}; processing speed; attention; cognitive flexibility; verbal fluency & Linear regression, correlation analyses & Higher \gls{brainPAD} associated with lower cognitive performance across multiple domains & Cross sectional; no causal inference; limited sample \\

\end{tabular}}
\end{table}
\clearpage
% ---------------------------------------------------------------------------
% Gaps in the Literature
\section{Gaps in the Literature}
First, most studies rely on \gls{TILDA} or similar cohorts of Irish and predominantly older adults (50+). This limits confidence in how well results transfer to other populations. The present project focuses on \gls{TILDA} to study ageing in depth but also makes use of the harmonised dataset, which links \gls{TILDA} with comparable UK and US cohorts. This helps place findings in a broader international context while retaining a core emphasis on older adults.

Second, there is an over reliance on physical health predictors such as comorbidity counts and mobility measures, while nutrition, social, religious or spiritual, and environmental variables are often overlooked. As shown in \cite{murphyPlasmaLuteinZeaxanthin2023}, nutritional biomarkers like lutein and zeaxanthin are strong predictors of frailty, yet these are rarely integrated into broader \gls{ML} pipelines. By contrast, studies such as \cite{bardonPredictorsIncidentMalnutrition2020} highlight the role of social support and functional decline in malnutrition risk. Together, these examples show that excluding nutrition and social factors risks producing models that reinforce the same biomedical biases seen in regression heavy work. From a Bayesian viewpoint, the omission of key features reduces the explanatory power of the model and weakens the posterior probability $P(\text{Outcome} \mid \text{All Features})$ compared to $P(\text{Outcome} \mid \text{Limited Features})$. Including a richer set of features increases the likelihood that the model accurately reflects the true data generating process.

Third, methodological choices remain conservative. Regression models still dominate, while more advanced approaches such as deep neural networks or transformers are rarely explored. For instance, \cite{matthewsPredictingMortalityIrish2022} achieved an \gls{AUC} of 0.78 using logistic regression for mortality prediction in \gls{TILDA}, whereas \cite{xuTailoredMachineLearning2023} reported an \gls{AUC} of 0.854 with ensemble methods for diabetes incidence. This contrast demonstrates both the dominance of regression and its weaker performance compared to flexible \gls{ML} models. Where ensemble or brain age approaches have been attempted, validation is almost always internal, with little cross cohort testing. Weak validation and limited feature sets reduce the generalisability of the estimated conditional probabilities and increase the risk of overfitting. Maximum likelihood estimation and Bayesian inference can provide a formal framework for assessing model fit and updating predictions as more diverse and representative data become available.

Fourth, many outcomes remain underexplored. Studies on dementia or frailty frequently focus on cognitive or physical decline while leaving aside outcomes critical to older adults, such as pain, sleep quality, and subjective quality of life. For example, \cite{murphyPlasmaLuteinZeaxanthin2023} used cross sectional plasma biomarkers to study frailty but did not assess sleep or pain, while gait speed models such as \cite{davisComparisonGaitSpeed2021} similarly neglect subjective quality of life. This shows how reliance on narrow outcomes leaves broader aspects of wellbeing unmeasured.

Fifth, cross sectional bias and data handling weaken reliability. Much nutritional work, such as \cite{murphyPlasmaLuteinZeaxanthin2023}, relies on single time point biomarker data, which cannot capture longitudinal change. By contrast, studies applying multi wave \gls{TILDA} data \cite{romero-ortunoTransitionsFrailtyPhenotype2021} demonstrate the added value of modelling temporal dynamics. Similarly, missing data are often handled through list wise deletion, as in \cite{matthewsPredictingMortalityIrish2022}, which risks biasing results toward healthier participants, whereas more robust imputation approaches remain underused. Missing or poorly handled data can distort the likelihood function and hence the posterior predictions, reducing confidence in the model's outputs.
 
Finally, interpretability challenges persist. While tools like \gls{SHAP} have been applied to highlight feature importance \cite{xuTailoredMachineLearning2023,davisLinearRegressionbasedMachine2021}, their use is inconsistent, and consensus on best practice has yet to emerge. Without transparent explanations, even accurate models may face barriers to clinical acceptance. Taken together, these examples show that ageing research continues to rely on narrow predictors, conservative models, and weak validation. Addressing these gaps requires integrative, diverse, and transparent approaches that link multidimensional predictors with explainable \gls{ML}.

% ---------------------------------------------------------------------------
% Summary
\section{Chapter Summary}
This chapter reviewed evidence on ageing and chronic disease across three domains, frailty and physical function, mental health, and nutrition and biological ageing. We also examined how machine learning has been applied to predict later life outcomes. 
The findings showed that while frailty is partly reversible, it remains strongly tied to mortality; that mental health both shapes and is shaped by physical decline; and that nutrition interacts with biological, metabolic, and social systems in ways that are often underestimated. 

While machine learning has improved prediction accuracy in ageing research, challenges remain in integrating diverse predictors and ensuring interpretability and longitudinal validation.

This study builds on previous research by leveraging \gls{TILDA} alongside the harmonised dataset linking UK and US cohorts, enabling longitudinal prediction of ageing outcomes. 
The main contribution lies in integrating nutrition, lifestyle, and clinical predictors, applying temporal modelling, and emphasising interpretability, addressing key limitations identified in prior work. 
These directions motivate the methodological choices outlined in Chapter~\ref{chap:Methodology}.



% ---------------------------------------------------------------------------
%: ----------------------- end of thesis sub-document ------------------------
% ---------------------------------------------------------------------------

